Enhanced Global PTM Discovery (G-PTM-D) with MetaMorpheus

### 2017-11-28 paper.tex
ajskf

Correct identification of protein post-translational modifications (PTMs) is crucial to understanding many aspects of protein function in biological processes.
G-PTM-D is a recently developed technique for global identification and localization of PTMs.
Spectral file calibration prior to applying G-PTM-D, and algorithmic enhancements in the peptide database search shown here significantly increase the accuracy and speed of PTM identification.
We enhance G-PTM-D by using multi-notch searches, and demonstrate its effectiveness in identification of numerous types of PTMs, including high-mass modifications such as glycosylations.
The changes described in this work lead to a 20\% increase in the number of identified modifications and an order of magnitude decrease in search time.
The complete workflow is implemented in MetaMorpheus, a software tool which integrates the database search procedure, identification of co-isolated peptides, spectral calibration and the enhanced G-PTM-D workflow.
Multi-notch searches are also shown to be useful in contexts other than G-PTM-D, producing superior results when used instead of standard narrow-window and open database searches.
\end{abstract}

@@

Numerous procedures for identification and localization of PTMs from "bottom-up" tandem mass spectrometric datasets exist.
These primarily address common modifications (e.g. phosphorylations) in well-studied systems, but global discovery tools, such as MODa\citep{Na_2011} and G-PTM-D\citep{Li_2016}, for identification and localization of a wide variety of different PTMs are also available.
G-PTM-D was developed in our laboratory and consists of three stages: 1) A wide precursor mass tolerance database search ('open' search)\citep{Chick_2015, Na_2011}, which provides mass differences between the identified peptides and the observed precursor peptide masses (hypothesized to differ in mass due to the presence of a PTM);
2) For those peptides for which the mass difference corresponds to the mass of a known PTM, a database augmentation step adds plausible localized PTMs to the corresponding protein entries in the search database;
3) A final standard tight precursor tolerance search ('narrow-window' search) using the augmented database to identify both modified and unmodified peptides subject to a standard false discovery rate (FDR) threshold based upon the target:decoy strategy\citep{Elias_2007}.

While G-PTM-D delivers confident identification of numerous types of PTMs, it has a few significant limitations.
The initial open search is slow, taking hours for modest size datasets.
Thus, high mass modifications such as glycosylations are out of reach, since widening the search window from the suggested $\pm 200$ Da increases the already-slow search time.
The final narrow-window search is also problematic, as it does not account for peptide-spectral matches (PSMs) with mass differences arising from misassignment of the monoisotopic peak during precursor deisotoping.
Discovery of modifications is limited to PTMs in the UniProt database: this is a significant limitation since that database does not include modifications arising from sample handling, and ignoring those modifications significantly degrades the final PTM identification rates.
Finally, PTMs that are similar in mass, such as phosphorylation (79.966331 Da) and sulfonation (79.956815 Da), are problematic since they are virtually indistinguishable in unprocessed spectral files.
These limitations motivated the work described here.

The enhanced G-PTM-D workflow is presented in Figure~\ref{fgr:diagram}.
Spectral calibration is essential for distinguishing modifications with similar mass, and so we recommend running the G-PTM-D procedure on calibrated spectra files: for this purpose we implemented a calibration procedure in MetaMorpheus.
To address the long search times of the open search, we replace the search in the first stage by a specialized search called a multi-notch search that only allows specific mass differences which correspond to pre-selected masses (e.g. PTMs and amino acids).
This change improves the specificity of the search and significantly decreases search time.
The database augmentation step is similar to the original G-PTM-D procedure, but the modification list is significantly expanded to accommodate modifications such as chemical adducts, glycosylations, lipids, and other important modifications.
The final narrow-window search is also replaced by a limited multi-notch search that accommodates errors in the precursor mass deisotoping (+1.0029 and + 2.0054), and the capability for identifying co-isolated peptides, further increasing the number of confidently identified peptides and PTMs.

@@

We developed a modified version of the Morpheus software for bottom-up spectral database searching\citep{Wenger_2013} that we call MetaMorpheus, which integrates the database search procedure with spectral calibration and the enhanced G-PTM-D workflow.
MetaMorpheus is freely available at \url{https://github.com/smith-chem-wisc/MetaMorpheus} as standalone Windows and Linux programs.
MetaMorpheus heavily utilizes an auxiliary library called mzLib, which is a vast library of software tools for analyzing and processing mass spectral data.x
mzLib can be found at \url{https://github.com/smith-chem-wisc/mzLib}.

@@

This work presents four major changes to the previously published G-PTM-D workflow:
a spectral calibration;
a replacement of the open search with a custom multi-notch search for PTM discovery;
replacement of the final traditional narrow mass search with a limited multi-notch search that permits exact matches even when incorrectly deisotoped precursor masses are present;
and lastly, addition of capacity for identification of multiple co-fragmented peptides from a single MS/MS scan.
To clarify the advantage of each change, they are first discussed separately, followed by an analysis of their combined effect upon overall G-PTM-D performance.

@@

We demonstrate the effectiveness of the calibration approach using a single Jurkat fraction (16) as an example.
Figure~\ref{fgr:fig1-calibErrors} displays the errors in m/z values of MS1 peaks identified by Step 1 of the calibration, both before and after the calibration procedure is applied.
These errors are plotted versus two important variables: the m/z value of the peak and the retention time of the corresponding scan.
In addition to these two parameters, the calibration function native to MetaMorpheus also considers peak intensity, injection time, total ion count, and other variables as potentially influencing the m/z errors.
Calibration successfully removes both linear and non-linear dependencies between the m/z errors and these variables.
Prior to calibration, 95\% of the errors fell between -8.4 and 3.1 ppm, and after calibration 95\% of the errors fell between -1.7 and 2.0 ppm.
MetaMorpheus calibrates both MS1 and MS2 scans (using MS2 fragment matches), but only the results from MS1 are shown here.
MetaMorpheus calibrated files are verified to be compatible with a variety of tools and workflows, including TPP\citep{Deutsch2010}, and MSFragger\citep{Kong_2017}.
In addition to calibrated spectra files, the calibration procedure outputs the standard deviations and averages of the errors post-calibration, which could then be used as search parameters; this is an alternative to using dedicated tools such as Param-Medic\citep{May2017} for determining the best search parameters for a given spectra file.

@@

The traditional narrow-window search limits the range of acceptable mass differences between a candidate peptide and the observed experimental precursor in order to find an exact match to the observed peptide in a protein database.
A mass tolerance of <20ppm is commonly used on high-resolution MS1 data, resulting in a highly restrictive search.
It is not uncommon for the precursor mass tolerance of narrow-window searches to be relaxed to as much as 2 or 3 Da.
One rationale for this is that errors in the deisotoping of MS1 spectra frequently result in a reported experimental precursor mass being different by 1.0029 or 2.0054 Da from the true isolated peptide mass.
The relaxed tolerance approach identifies those peptides correctly.
A second rationale for this approach is that multiple peptides are frequently co-isolated and co-fragmented with all peptide fragments being observed in the same tandem mass spectrum.
Frequently, the best-fragmented peptide is different from the one corresponding to the precursor mass reported by the instrument, and the relaxed tolerance approach may identify this peptide, provided the reported charge state matches.
The major limitation of this approach is that a 2 or 3 Da tolerance includes mass shifts that, when present, may indicate that the matched peptide is not the correct identification, arguably defeating the purpose of a narrow-window search.
This may happen when the mass difference arises due to the identified peptide being different from the actual isolated precursor by a low mass modification or amino acid substitutions.
The described issues are apparent in mass difference histograms in Figure~\ref{fig:fig4jurkat-1012}, generated by compiling the results of a 2.5 Da tolerance narrow-window search.
These plots clearly show that numerous peptides have missed monoisotopic (Mm) peaks, deamidation, or an amino acid substitution.

@@

To illustrate the described advantages, we compared the three types of searches using all 28 calibrated Jurkat fractions (Table~\ref{tbl:singleVsMultiNotch}).
The limited notch search yields the largest number of confidently identified PSMs with only a small fraction being misassigned.
We consider PSMs to be misassigned if the G-PTM-D procedure determines a higher-scoring modified peptide for a PSM (see the last section for details).
One example of a misassigned PSM, in the 2.5 Da Search, is the example where the correct identification has oxidation and ammonia loss, but was identified as an unmodified peptide with score 9.039 and mass difference of -1.03335 Da, with sequence \texttt{QAFDDAIAELDTLNEDSYK}, in Jurkat Frac9 scan 20962.
With G-PTM-D added modifications, the correct peptide is identified with sequence \texttt{[Ammonia los]QAFDDAIAELDTLNEDSY[Oxidation]K}. It has a mass difference of -0.00171(-0.79458 ppm) and a score of 19.222.
While this kind of misassignment is not possible in the 5ppm narrow-window search, such missed monoisotopes are often assigned to a wrong sequence.
This explains the higher number of misassignments in the 5ppm search over the number in the multi-notch search.
Finally, even the multi-notch search has some misassignments.
One example, in Jurkat Frac6 scan 15602, is the wrong identification of \texttt{SIQEIQELDK[acetyl]DDESLR} with a score of 10.153, when the correct identification, after the G-PTM-D procedure, is \texttt{S[acetyl]IQEIQELDKDDESLR} with score 16.249, where the masses are identical.
While all of the described searches have some misassigned false positive identifications, the multi-notch search minimizes them while also yielding the most PSMs.

@@

The alternative we recommend, is the use of a multi-notch search, which offers a means of reducing the search space and limiting the number of theoretical candidates to those plausible for the scan, increasing the search speed and most importantly, producing the most correct PSMs with the fewest false positive PSM identifications.
We compared searches of all Jurkat fractions using a $\pm 500$ Da wide mass search to a multi-notch search with notches at 0 and 79.961 with 0.02 Dalton tolerance.
The advantage of the multi-notch search over the open search is shown in Table~\ref{tbl:multiVsWide} by at least a $10\%$ increase in the number of target PSMs with score of at least 9.079, for both regions (0 Da no-PTM peptides and +79.96 Da phosphorylated and sulfonated peptides).
It is important to point out that the difference in the numbers of PSMs is not due to FDR considerations, but rather to erroneous peak assignments in open searches, that misassign PSMs due to the extreme freedom in precursor masses allowed in open searches.
A specific example of a PSM that has been erroneously assigned in a open search is scan 12835 in Jurkat frac 5.
The open search identifies \texttt{ATPARAPESPPSADPALVAGPAEEAEC[Carbamidomethyl]PPPR} as the best match with a score of 17.231 and mass difference of -416.31028, and a single missed cleavage.
The better match is identified by the multi-notch search, as \texttt{APESPPSADPALVAGPAEEAEC[Carbamidomethyl]PPPR} with a score of 16.246 and mass difference of 79.9655, and no missed cleavages.
So, instead of having a PSM with an unidentifiable mass shift of -416.31028 (mass of removal of \texttt{ATPAR} and addition of phosphorylation), we have an easily identifiable mass shift corresponding to a phosphorylation site.

@@

Consider the Jurkat Frac 12 scan 12336 with precursor scan 12333.
Isolation m/z is 726.65 Th and isolation width is 2.5 Th.
It is shown in Figure~\ref{fig:fig5-coIsolationSpectrum}.
There are at least three different peptides being co-isolated and co-fragmented, and there is evidence in the MS/MS spectrum for all of them.

@@

\begin{table}[]
\centering
\caption{Effects of Replacing Initial Open Search by a Multi-Notch Search}
\label{my-labelff}
\begin{tabular}{ll|l|l}
                      &        & Open Search & Multi-Notch Search\\
\multirow{2}{*}{Time} & Jurkat & 22.1 hrs         & 2.9 hrs    \\
                      & Mouse  & 13.5 hrs         & 1.2 hrs   \\
\end{tabular}
\end{table}

After the final search using the augmented database is conducted (narrow-window in the original, and limited multi-notch in the enhanced G-PTM-D), the numbers of PSMs, peptides, and proteins all increase as shown in Table~\ref{tab:table2}.
We identified hundreds of glycosylated peptides in these unenriched samples, with many of these modifications exceeding 1000 Da.
These enhancements demonstrate that the enhanced G-PTM-D approach provides great value to bottom-up proteomic data analysis.
It is fast enough included as part of every regular work-flow, and will yield significantly increased numbers of correctly identified PTM peptides.

@@

>>>>

Correct identification of protein post-translational modifications (PTMs) is crucial to understanding many aspects of protein function in biological processes.
G-PTM-D is a recently developed technique for global identification and localization of PTMs.
Spectral file calibration prior to applying G-PTM-D, and algorithmic enhancements in the peptide database search shown here significantly increase the accuracy, speed and scope of PTM identification.
We enhance G-PTM-D by using multi-notch searches, and demonstrate its effectiveness in identification of numerous types of PTMs, including high-mass modifications such as glycosylations.
The changes described in this work lead to a 20\% increase in the number of identified modifications and an order of magnitude decrease in search time.
The complete workflow is implemented in MetaMorpheus, a software tool which integrates the database search procedure, identification of co-isolated peptides, spectral calibration and the enhanced G-PTM-D workflow.
Multi-notch searches are also shown to be useful in contexts other than G-PTM-D, producing superior results when used instead of standard narrow-window and open database searches.
\end{abstract}

@@

Numerous procedures for identification and localization of PTMs from "bottom-up" tandem mass spectrometric datasets exist.
These primarily address common modifications (e.g. phosphorylations) in well-studied systems, but global discovery tools, such as MODa\citep{Na_2011} and G-PTM-D\citep{Li_2016}, for identification and localization of a wide variety of different PTMs are also available.
G-PTM-D was developed in our laboratory and consists of three stages: 1) A wide precursor mass tolerance database search ('open' search)\citep{Chick_2015, Na_2011}, which provides mass differences between the identified peptides and the observed precursor peptide masses, hypothesized to differ in mass due to the presence of a PTM;
2) For those peptides for which the mass difference corresponds to the mass of a known PTM, a database augmentation step adds plausible localized PTMs to the corresponding protein entries in the search database;
3) A final standard tight precursor tolerance search ('narrow-window' search) using the augmented database to identify both modified and unmodified peptides subject to a standard false discovery rate (FDR) threshold based upon the target:decoy strategy\citep{Elias_2007}.

While G-PTM-D delivers confident identification of numerous types of PTMs, it has a few significant limitations.
The initial open search is slow, taking hours for modest size datasets.
Thus, high mass modifications such as glycosylations are out of reach, since widening the search window from the suggested $\pm 200$ Da increases the already-slow search time.
The final narrow-window search is also problematic, as it does not account for peptide-spectral matches (PSMs) with mass differences arising from misassignment of the monoisotopic peak during precursor deisotoping.
Discovery of modifications is limited to PTMs in the UniProt database: this is a significant limitation since that database does not include modifications arising from sample handling, and ignoring those modifications significantly degrades the final PTM identification rates.
Finally, PTMs that are similar in mass, such as phosphorylation (79.9663 Da) and sulfonation (79.9568 Da), are problematic since they are virtually indistinguishable in unprocessed spectral files.
These limitations motivated the work described here.

The enhanced G-PTM-D workflow is presented in Figure~\ref{fgr:diagram}.
Spectral calibration is essential for distinguishing modifications with similar mass, and so we recommend running the G-PTM-D procedure on calibrated spectra files: for this purpose we implemented a calibration procedure in MetaMorpheus.
To address the long search times of the open search, we replace the search in the first stage by a specialized search called a multi-notch search that only allows specific mass differences which correspond to pre-selected masses (e.g. PTMs and amino acids).
This change improves the specificity of the search and significantly decreases search time.
The database augmentation step is similar to the original G-PTM-D procedure, but the modification list is significantly expanded to accommodate modifications such as chemical adducts, glycosylations, lipids, and other important modifications.
The final narrow-window search is also replaced by a limited multi-notch search that accommodates errors in the precursor mass deisotoping (+1.0029 and +2.0054), and the capability for identifying co-isolated peptides, further increasing the number of confidently identified peptides and PTMs.

@@

We developed a modified version of the Morpheus software for bottom-up spectral database searching\citep{Wenger_2013} that we call MetaMorpheus, which integrates the database search procedure with spectral calibration and the enhanced G-PTM-D workflow.
MetaMorpheus is freely available at \url{https://github.com/smith-chem-wisc/MetaMorpheus}.
MetaMorpheus heavily utilizes an auxiliary library called mzLib, which is a vast library of software tools for analyzing and processing mass spectral data.
mzLib can be found at \url{https://github.com/smith-chem-wisc/mzLib}.

@@

This work presents four major changes to the previously published G-PTM-D workflow:
a spectral calibration;
a replacement of the open search with a custom multi-notch search for PTM discovery;
replacement of the final traditional narrow-window search with a limited multi-notch search that permits exact matches even when incorrectly deisotoped precursor masses are present;
and lastly, addition of capacity for identification of multiple co-fragmented peptides from a single MS/MS scan.
To clarify the advantage of each change, they are first discussed separately, followed by an analysis of their combined effect upon overall G-PTM-D performance.

@@

We demonstrate the effectiveness of the calibration approach using Jurkat fraction 16 as an example.
Figure~\ref{fgr:fig1-calibErrors} displays the errors in m/z values of MS1 peaks identified by Step 1 of the calibration, both before and after the calibration procedure is applied.
These errors are plotted versus two important variables: the m/z value of the peak and the retention time of the corresponding scan.
In addition to these two parameters, the calibration function native to MetaMorpheus also considers peak intensity, injection time, total ion count, and other variables as potentially influencing the m/z errors.
Calibration successfully removes both linear and non-linear dependencies between the m/z errors and these variables.
Prior to calibration, 95\% of the errors fell between -8.4 and 3.1 ppm, and after calibration 95\% of the errors fell between -1.7 and 2.0 ppm.
MetaMorpheus calibrates both MS1 and MS2 scans (using MS2 fragment matches), but only the results from MS1 are shown here.
MetaMorpheus calibrated files are verified to be compatible with a variety of tools and workflows, including TPP\citep{Deutsch2010}, and MSFragger\citep{Kong_2017}.
In addition to calibrated spectra files, the calibration procedure outputs the standard deviations and averages of the errors post-calibration, which could then be used as search parameters; this is an alternative to using dedicated tools such as Param-Medic\citep{May2017} for determining the best search parameters for a given spectra file.

@@

The traditional narrow-window search limits the range of acceptable mass differences between a candidate peptide and the observed experimental precursor in order to find an exact match to the observed peptide in a protein database.
A mass tolerance of <20ppm is commonly used on high-resolution MS1 data, resulting in a highly restrictive search.
It is not uncommon for the precursor mass tolerance of narrow-window searches to be relaxed to as much as 2 or 3 Da.
One rationale for this is that errors in the deisotoping of MS1 spectra frequently result in a reported experimental precursor mass being different by 1.0029 or 2.0054 Da from the true isolated peptide mass.
The relaxed tolerance approach identifies those peptides correctly.
A second rationale for this approach is that multiple peptides are frequently co-isolated and co-fragmented with all peptide fragments being observed in the same tandem mass spectrum.
Frequently, the best-fragmented peptide is different from the one corresponding to the precursor mass reported by the instrument, and the relaxed tolerance approach may identify this peptide, provided the reported charge state matches.
The major limitation of this approach is that a 2 or 3 Da tolerance includes mass shifts that, when present, may indicate that the matched peptide is not the correct identification, arguably defeating the purpose of a narrow-window search.
This may happen when the mass difference arises due to the identified peptide being different from the actual isolated precursor by a low mass modification or amino acid substitutions.
The described issues are apparent in mass difference histograms in Figure~\ref{fig:fig4jurkat-1012}, generated by compiling the results of a 2.5 Da tolerance narrow-window search.
These plots clearly show that numerous peptides have missed monoisotopic peaks, deamidation, or an amino acid substitution.

@@

To illustrate the described advantages, we compared the three types of searches using all 28 calibrated Jurkat fractions (Table~\ref{tbl:singleVsMultiNotch}).
The limited notch search yields the largest number of confidently identified PSMs with only a small fraction being misassigned.
We consider PSMs to be misassigned if the G-PTM-D procedure determines a higher-scoring modified peptide for a PSM (see the last section for details).
One example of a misassigned PSM, in the 2.5 Da Search, is the example where the correct identification has oxidation and ammonia loss, but was identified as an unmodified peptide with score 9.039 and mass difference of -1.03335 Da, with sequence \texttt{QAFDDAIAELDTLNEDSYK}, in Jurkat fraction 9 scan 20962.
With G-PTM-D added modifications, the correct peptide is identified with sequence \texttt{[Ammonia los]QAFDDAIAELDTLNEDSY[Oxidation]K}. It has a mass difference of -0.00171(-0.79458 ppm) and a score of 19.222.
While this kind of misassignment is not possible in the 5ppm narrow-window search, such missed monoisotopes are often assigned to a wrong sequence.
This explains the higher number of misassignments in the 5ppm search over the number in the multi-notch search.
Finally, even the multi-notch search has some misassignments.
One example, in Jurkat fraction 6 scan 15602, is the wrong identification of \texttt{SIQEIQELDK[acetyl]DDESLR} with a score of 10.153, when the correct identification, after the G-PTM-D procedure, is \texttt{S[acetyl]IQEIQELDKDDESLR} with score 16.249, where the masses are identical.
While all of the described searches have some misassigned false positive identifications, the multi-notch search minimizes them while also yielding the most PSMs.

@@

The alternative we recommend, is the use of a multi-notch search, which offers a means of reducing the search space and limiting the number of theoretical candidates to those plausible for the scan, increasing the search speed and most importantly, producing the most correct PSMs with the fewest false positive PSM identifications.
We compared searches of all Jurkat fractions using a $\pm 500$ Da wide mass search to a multi-notch search with notches at 0 and 79.961 with 0.02 Dalton tolerance.
The advantage of the multi-notch search over the open search is shown in Table~\ref{tbl:multiVsWide} by at least a $10\%$ increase in the number of target PSMs with score of at least 9.079, for both regions (0 Da no-PTM peptides and +79.96 Da phosphorylated and sulfonated peptides).
It is important to point out that the difference in the numbers of PSMs is not due to FDR considerations, but rather to erroneous peak assignments in open searches, that misassign PSMs due to the extreme freedom in precursor masses allowed in open searches.
A specific example of a PSM that has been erroneously assigned in a open search is scan 12835 in Jurkat fraction 5.
The open search identifies \texttt{ATPARAPESPPSADPALVAGPAEEAEC[Carbamidomethyl]PPPR} as the best match with a score of 17.231 and mass difference of -416.31028, and a single missed cleavage.
The better match is identified by the multi-notch search, as \texttt{APESPPSADPALVAGPAEEAEC[Carbamidomethyl]PPPR} with a score of 16.246 and mass difference of 79.9655, and no missed cleavages.
So, instead of having a PSM with an unidentifiable mass shift of -416.31028 (mass of removal of \texttt{ATPAR} and addition of phosphorylation), we have an easily identifiable mass shift corresponding to a phosphorylation site.

@@

Consider the Jurkat fraction 12 scan 12336 with precursor scan 12333.
Isolation m/z is 726.65 Th and isolation width is 2.5 Th.
It is shown in Figure~\ref{fig:fig5-coIsolationSpectrum}.
There are at least three different peptides being co-isolated and co-fragmented, and there is evidence in the MS/MS spectrum for all of them.

@@

\begin{table}[]
\centering
\caption{Effects of Replacing Initial Open Search by a Multi-Notch Search}
\label{my-labelff}
\begin{tabular}{llll}
                      &        & Open Search & Multi-Notch Search\\
\multirow{2}{*}{Time} & Jurkat & 22.1 hrs         & 2.9 hrs    \\
                      & Mouse  & 13.5 hrs         & 1.2 hrs   \\
\end{tabular}
\end{table}

After the final search using the augmented database is conducted (narrow-window in the original, and limited multi-notch in the enhanced G-PTM-D), the numbers of PSMs, peptides, and proteins all increase as shown in Table~\ref{tab:table2}.
We identified hundreds of glycosylated peptides in these unenriched samples, with many of these modifications exceeding 1000 Da, see Table~\ref{tab:table3}.
These enhancements demonstrate that the enhanced G-PTM-D approach provides great value to bottom-up proteomic data analysis.
It is fast enough included as part of every regular work-flow, and will yield significantly increased numbers of correctly identified PTM peptides.

@@

\begin{table}[]
\centering
\caption{Highest mass PTMs identified}
\label{tab:table3}
\begin{tabular}{ll}
Mass (Daltons)&	Mod\\
1378.474&	Hex(6)HexNAc(2)\\
1216.4236&	Hex(5)HexNAc(2)\\
1054.3719&	Hex(4)HexNAc(2)\\
1038.375&	dHex(1)Hex(3)HexNAc(2)\\
963.3177&	Hex(1)HexNAc(1)NeuAc(1)NeuGc(1)
\end{tabular}
\end{table}
